---
title: "Tuning TROLL Workflow"
output: 
  rmarkdown::html_vignette:
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Tuning TROLL Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
body {
  max-width: 90% !important;  /* Or 1200px if you want a fixed limit */
  margin: auto;               /* Centers the page */
  padding: 1em;               /* Prevents text from hitting the edge */
 font-size: 18px;            /* Increases the general text size */
  line-height: 1.6;           /* Adds space between lines for readability */
}

/* Optional: If you want your code blocks to be larger too */
code, pre {
  font-size: 16px !important; 
}

/* Ensure code blocks and images also expand */
pre, img {
  max-width: 100%;
}
</style>

```{r Setup, include=FALSE}
library(ImportUtils)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::knit_hooks$set(error = function(x, options) {
  paste0("<pre style='color: red; background-color: #fee;'><code>", x, "</code></pre>")
})

file_location_path <- system.file(
  "extdata",
  "TROLL_exdata.csv",
  package = "ImportUtils"
)
```
## Overview

The AquaTROLL utility functions in `ImportUtils` provide tools for processing electronic 
logs generated by the VuSitu app. The primary workhorse is `TROLL_profile_compiler()`,
which wraps the functionality of the supporting functions into a single interface for batch processing.
Ideally, this is the only function required to process raw AquaTroll data (see the *TROLL-QuickStart* vignette). 

In the event that an error is triggered, or output is not as expected, each component function used within 
`TROLL_profile_compiler()` is also exported individually, allowing users to inspect intermediate steps,
troubleshoot issues, and better understand how the data are transformed throughout the workflow. 

The workflow is:
  
1) Read Raw Data (`TROLL_read_data()`)
2) Standardize column names and required data (`TROLL_rename_cols()`)
3) Detect and flag stationary periods in sonde data (`is_stationary()`)
4) Detect and flag stable data by sensor type (`TROLL_sensor_stable()`)
5) Summarize stationary + stable data (`TROLL_stable_summary()`)

These functions assume that raw VuSitu `.html` files have already been converted to `.csv` format. 
This vignette walks through a complete example workflow using the individual functions, followed 
by the wrapper function. The goal is to demonstrate the purpose of each step and help users 
make informed decisions for efficient and accurate data processing.

## Step 1: Read Raw Data

The `TROLL_read_data()` function reads raw `.csv` files exported from VuSitu. It uses pattern matching 
to locate the "Date Time" column and determine where the data begin.

The function checks whether seconds are included in the datetime values and attempts to standardize 
formatting if needed. Because the "Date Time" column is used to identify the start of the dataset, 
it must be present in the input file.

```{r Data Read, echo=TRUE, warning=FALSE, include=TRUE}
dat_read <- TROLL_read_data(path = file_location_path)
```

At this stage, the data are imported without interpretation or modification beyond basic formatting (e.g
`Date Time` and `Depth (m)` column formatting are enforced).  

## Step 2: Standardize Column Names

Downstream functions rely on consistent column naming; this function enforces standard (canonical)
naming conventions via `troll_column_dictionary`. 

```{r Rename Columns, echo=TRUE,include=TRUE}
dat_rename <- TROLL_rename_cols(
  df = dat_read,
  colname_dictionary = troll_column_dictionary
)
```
### If you see an error about "Unknown Columns ..." 

Column renaming is achieved by pattern matching. If unknown columns are detected **an error** will be displayed
indicating that the dictionary should be updated if these columns represent new sensor data. However, it is 
possible to work around missing names by creating a unique dictionary object and passing it to the `colname_dictionary`
argument using `troll_column_dictionary` as a template and adding additional rows as required. 

```{r Column Dictionary,echo=TRUE,include=TRUE, error=TRUE}
# Drop "pH_units" from the dictionary to simulate a missing parameter
missing_dictionary_row <- troll_column_dictionary[-5,] 

# With "pH_units" missing, an ERROR is shown:
dat_rename <- TROLL_rename_cols(
  df = dat_read,
  colname_dictionary = missing_dictionary_row
)

# Create a new row with the missing information, formatted appropriately
new_row <- troll_column_dictionary[5,]
new_row

# Add to the dictionary object
good_dictionary <- missing_dictionary_row |>
  dplyr::add_row(new_row)

# Apply updated dictionary
dat_rename <- TROLL_rename_cols(
  df = dat_read,
  colname_dictionary = good_dictionary
)

names(dat_rename)

```

Columns associated with the TROLL COMM are automatically detected using an internal dataset of TROLL COMM serial numbers.
The TROLL COMM temperature column is renamed to avoid conflicts with water temperature. COMM columns are stripped away with 
metadata whent `strip_metadata` = TRUE. 

The `trollcomm_serials` arguments allows users to use a custom vector of serial numbers associated with 
TROLL COMMunits, if the internal dataset is out of date. 

```{r COMM Serials,echo=TRUE,include=FALSE}
TROLL_rename_cols(
  df = dat_read,
  trollcomm_serials = c(1151975, 1153542) # Add additional serial #'s here
)
```
Metadata columns (e.g. "latitude", "longitude", "battery" etc.) are removed by default, but can be retained by toggling
`strip_metadata` == FALSE. The inclusive list of metadata columns are defined within the column dictionary:
  
```{r Metadata,echo=TRUE,include=TRUE}
troll_column_dictionary[troll_column_dictionary$meta,]

dat_meta <- TROLL_rename_cols(
  df = dat_read,
  strip_metadata = FALSE # Retain metadata columns
)
```

For troubleshooting, the `verbose` argument may be toggled to allow printing of messages to the console regarding 
presence/absense of Troll-Comm columns, and which parameters were detected. 

```{r Verbose,echo=TRUE,include=TRUE}
dat_echo <- TROLL_rename_cols(
  df = dat_read,
  verbose = TRUE
)
```
## Step 3: Detect Stationary Periods in Sonde Depth Data
Once naming schema are standardized, the next step is to detect periods when the sonde was stationary in the water column. 
**You can simply pass a path to the `df` argument of `is_stationary` and steps 1 & 2 will be completed within the function**
The `is_stationary` function detects periods when the sonde is stationary based on a rolling range, and assigns
`is_stationary_status` based on the amount of time the sonde has remained within a `depth_range_threshold`. 

### How to Tune Stationary Detection

There are three parameters that control how stationary periods are identified: 

* `depth_range_threshold`
Controls how much vertical movement is allowed within the rolling window. 
  * smaller = stricter 
  * larger = more persmissive (risk of false positives)
* `stationary_secs`
Minimum duration required for a depth to be considered *fully stationary* (999)
  * Larger values = only long duration pauses retained
  * Smaller values = shorter pauses included
* `rolling_range_secs`
Length of the backward-looking window used to comput depth range
  * Larger values = smoother, more conservative detection
  * Smaller values = more sensitive to short-term variation

```{r First Stationary, echo=TRUE, include=TRUE,fig.align='center', fig.width=8, fig.height=6, warning=FALSE}
dat_stationary <- 
is_stationary(df = dat_rename,
              depth_range_threshold = 0.025, # Default = 0.05
              stationary_secs = 40,          # Default = 45
              rolling_range_secs = 10,       # Default = 10
              plot = TRUE)
```
**Figure 1**: Plot output from `is_stationary` indicating which stationary depths were extracted (horizontal dashed lines
in top panel). 

Once stationary periods are detected, `is_stationary` appends `stationary_depth`as the mean of the stationary observation depths 
to the input `df`. As you can see in the legend, stationary periods longer than `stationary_secs` are flagged with **999**, 
indicating they are fully stationary. This is critical for downstream use, **only `is_stationary_status` 999 observations are 
considered for stability detection**. 

In this example, we can see in the bottom panel of **Figure 1**, that the `depth_range_threshold` is too strict, causing splitting 
of blocks that hover near the same depth. 

```{r stationary adjust range, echo=TRUE, include=TRUE,fig.align='center', fig.width=8, fig.height=6, warning=FALSE}
dat_stationary <- 
is_stationary(df = dat_rename,
              depth_range_threshold = 0.05, # increased to avoid erroneous splits (this is the default value)
              stationary_secs = 40,
              rolling_range_secs = 10,
              plot = TRUE)
```
**Figure 2**: Plot output from `is_stationary` after adjustment to `depth_range_threshold`

Now, the blocks are continuous, but it is clear that if we want to include data from > 5 m, we will have to relax `stationary_secs` to
allow shorter periods through. If we use `stationary_secs` = 24, all blocks to ~ 8.0 m will be included. 

```{r stationary adjust time, echo=TRUE, include=TRUE,fig.align='center', fig.width=8, fig.height=6, warning=TRUE}
dat_stationary <- 
is_stationary(df = dat_rename,
              depth_range_threshold = 0.05, # increased to avoid erroneous splits (this is the default value)
              stationary_secs = 24,
              plot = TRUE)
```
**Figure 3**: Plot output form `is_stationary` after adjustment of `stationary_secs`. 

**The "Warning: Inconsistent sampling intervals..."** is created by an internal helper that extracts sampling interval from 
the DateTime column. What it's telling you is that there are inconsistent temporal "jumps" in the data, usually caused by the Troll-Comm 
glitching. This is only an issue if the "Dominant Interval" is not detected as what it should be (2 seconds in this example is correct).
Stationary periods are calculated based on time, so large jumps are not a problem as long as the sonde was held in the same position. 

### Now we have flagged stationary depth from 0 - 8 meters at ~ 1.0 m intervals. 

## Step 4: Detect Stable Sensor Data Inside Stationary Blocks

The `TROLL_sensor_stable()` function handles one column of sensor data at a time. It is iteratively applied 
across all parameter columns within `TROLL_profile_compiler` using default range and slope threshold values (`stability_ranges`).
In this section, the function is applied outside of the compiler to demonstrate how tuning of `TROLL_sensor_stable` 
may be applied for individual parameters, if necessary. 

#### How Stability Detection Works 

At the core, stability detection functions the same across all data parameters: 

1. Stationary blocks of data are identified using `is_stationary`
2. An initial "settling period" is trimmed from the start of each block via `settling_secs`
3. Slope and range are calculated across each stationary block independently 
4. Observations are dropped iteratively from the start of each stationary block, slope and range are re-calculated on the remainder
until `min_secs` of time remains in the stationary block
5. Once `slope_thresh` AND `range_thresh` are met, the remainder of the block is used to calculate the median value

**Optical Probes (Turbidity, Chlorophyll, BGA)** utilize all "stationary" data after trimming of `stationary_secs` due to optical 
sensor behavior. A warning is shown if the slope threshold is exceeded for one of the optical parameters, allowing the user 
to explore if there are suspect data at the specific stationary depth output by `TROLL_stable_summary`. 

### Time Series Plots

It may not be intuitive at first to view water column profile data as time series plots, however, by doing so the user is able 
to visually assess how parameter values are changing through the cast relative to stationary blocks of data, and whether the 
extracted "stable" data points are representative of reality. 

```{r pH Stability, echo=TRUE, include=TRUE,fig.align='center', fig.width=8, fig.height=4}
dat_stable <- TROLL_sensor_stable(
  df = dat_stationary,
  value_col = pH_units,
  settling_secs = 10,  # Front end trimming of stationary blocks
  min_median_secs = 5, # Minimum amount of data retained to calculate median if stability is not detected
  slope_thresh = 0.05,
  range_thresh = 0.02,
  plot = TRUE           # Optional plotting toggled ON
)

```
**Figure 4**: Plot output from `TROLL_sensor_stable` for pH. 

The horizontal dashed lines represent the `range_thresh` value, which can be useful for tuning. The dark purple points are the extracted "stable"
data which are used to calculate the median at each stationary depth by `TROLL_stable_summary`. 

In **Figure 4** an exponential type relationship is apparent across each stationary block of data, with stability occurring late in each block,
if at all. 

Due to the nature of the parameters being measured, as well as how optical sensors record data, values from optical sensors (turbidity, 
chlorophyll, BGA) tend to be noisier and not exhibit such exponential changes (**Figure 5**). 

```{r Turb Stability, echo=TRUE, include=TRUE,fig.align='center', fig.width=8, fig.height=4}
dat_stable <- TROLL_sensor_stable(
  df = dat_stationary,
  value_col = turbidity_NTU,
  settling_secs = 10,  # Front end trimming of stationary blocks
  min_median_secs = 5, # Minimum amount of data retained to calculate median if stability is not detected
  slope_thresh = NULL, # Use defaults
  range_thresh = NULL, # Use defaults
  plot = TRUE          # Optional plotting toggled ON
)
```
**Figure 5**: Plot output from `TROLL_sensor_stable` for turbidity. 

The function here warns us that the slope threshold was exceeded at 3 stationary depths:

1. At 2.90 m (the fourth stationary block accounting for depth ~ 0) this is clearly caused by a small number of observations at the tail of the stationary block of ~ 2.4 - 2.5 NTU
2. At 6.83 m the data exhibits consistent drift in the negative direction
3. At 7.68 m is primarily the result of higher NTU values (on a relative basis such as percent change this would appear less impactful)

There are no range threshold lines in **Figure 5** because it is not used for optical sensors. 

## Step 5: Summarize Stable Data

With data flagged as stable in hand, the `TROLL_stable_summary` function compiles the output into a single value (default = median)
for each stationary depth. This is carried out internally in `TROLL_profile_compiler` when `summarize_data` is set to `TRUE`. For 
demonstration here we compile the summary data for turbidity used in **Figure 5**. 

```{r stable summary,echo=TRUE,include=TRUE}
dat_summary <- TROLL_stable_summary(
  df = dat_stable,
  group_col = stationary_depth
) |>
  dplyr::select(stationary_depth, turbidity_NTU)
dat_summary
```

## Step 6: Putting it all Together

The `TROLL_profile_compiler` function is a high level wrapper that automatically detects parameters using the `troll_colname_dictionary` 
and iterates over them with `TROLL_sensor_stable` to provide summarized data from an entire AquaTroll sonde cast. Inputs to the compiler
are named to indicate which of the other package functions they relate to: 

1. prefix "stn_" is associated with inputs to `is_stationary`
2. prefix "stbl_" is associated with input to `TROLL_sensor_stable`

The output from the compiler is either: 

1. A list of length (2) if `summarize_data` is set to `TRUE`

    a. The first list element is named "Flagged_Data", and is the raw input dataframe with a "<param>_stable" flag appended to each parameter column.

    b. The second list element is the summary output from `TROLL_stable_summary`.
  
Instead of timeseries plots, the compiler function outputs classical "profile" plots for each of the core parameters in the data when `plot = c(Final = TRUE)`. 

### Updating Inputs to the Compiler Based on Tuning Results

If the default values for any of the inputs needs to be tuned, they can updated: 

**Table 1**: Input naming conventions among package functions. 

|TROLL_profile_compiler|Other Package Function|Other Function Input|
|:---|:---|:---|
|stn_depthrange|is_stationary()|depth_range_threshold|
|stn_secs|is_stationary()|stationary_secs|
|stn_rollwindow_secs|is_stationary()|rolling_range_secs|
| | | |
|stbl_settle_secs|TROLL_sensor_stable()|settling_secs|
|stbl_min_secs|TROLL_sensor_stable()|min_median_secs|
|stbl_range_thresholds|TROLL_sensor_stable()|range_thresh|
 
Default values for individual `range_thresh` and `slope_thresh` by parameter are given by `stability_ranges`:

```{r Show Thresholds,echo=FALSE,include=TRUE}
stability_ranges
```

These may be updated by passing a named numeric vector. If we want to relax the `range_thresh` for dissolved oxygen and pH as an example: 

```{r custom range thresholds,echo=TRUE,include=TRUE, warning=FALSE}
# Create a named vector of new range threshold values
custom_ranges <- c('DO_mgL' = 0.2, 'pH_units' = 0.15)

dat_final <- TROLL_profile_compiler(
  path = file_location_path,
  stn_depthrange = 0.05,
  stn_secs = 24,                                   # Note use of value from tuning above
  stbl_range_thresholds = custom_ranges,           # Provide updated custom ranges
  summarize_data = TRUE,                           # Output will be a list w/ both flagged and summary data
  plot = c(Final = FALSE, Stationary = FALSE)
)
```

**Currently updating of slope thresholds is not allowed**

If `summarize_data` is `FALSE`, output will just return the raw dataframe with a "_stable" flag column for each parameter. 
`Flagged_Data` is always returned, but the `Summary_Data` can be extracted from the output list when `summarize_data = TRUE`: 

```{r summary table,echo=TRUE,include=TRUE}
flag_data <- dat_final$Flagged_Data
summary <- dat_final$Summary_Data
```
**Table 2**: Example of `$Summary_Data` from the compiler function output. 
```{r table 2, echo=FALSE,include=TRUE}
knitr::kable(summary)
```

The optional plotting should be used to visually ensure that each parameter was compiled as expected. 
The plots shown below are the classical view of profile data, with data values on the x axis and depth on the y-axis. 
Optional plotting is toggled with `plot = c(Final = TRUE)`, `plot = c(Final = TRUE, Stationary = TRUE)` (to show the 
stationary plot like **Figure 3**), or `plot = TRUE` (toggles both plots ON). 

If any individual parameter looks incorrect, it is recommended to move outside of `TROLL_profile_compiler` and 
work in the individual workflow to tune, and bring those values back to the wrapper. 

```{r final plots,echo=FALSE, include=TRUE,fig.align='center', fig.width=4, fig.height=7, warning=FALSE}
dat_final <- TROLL_profile_compiler(
  path = file_location_path,
  stn_depthrange = 0.05,
  stn_secs = 24,                                   # Note use of value from tuning above
  stbl_range_thresholds = custom_ranges,           # Provide updated custom ranges
  summarize_data = TRUE,                           # Output will be a list w/ both flagged and summary data
  plot = c(Final = TRUE, Stationary = FALSE)
)
```


